Supporting Dynamic Data Partitioning in Parallel Programs

### 2012-11-01 doc/types-pldi2013/intro.tex
intro

Our own experience writing high-performance 
using the current industry standard mix of MPI, shared-memory
threads, and CUDA, has taught us that a
single, static data partitioning is insufficient.  In practice, the
best way to partition data is often a function of the data
itself---i.e., the partitions must be dynamically computed.
Furthermore, applications often need multiple, simultaneous partitions of
the same data, providing different views on to that data.

In this work, we present the design of Legion \cite{Legion12},
a parallel programming model that efficiently supports multiple, dynamic partitions
of data. Legion's {\em logical regions} are an
abstraction capturing {\em locality} (data that will be used together,
and therefore should be colocated) and {\em independence} (disjoint data
that may be operated on in parallel).  Logical regions have 
properties that allow programmers to dynamically communicate information
about the structure of program data to Legion:
\begin{itemize}
\item  logical regions are first-class values in Legion
and may be dynamically allocated and stored in data structures;

@@

\item programmers give logical regions {\em privilege} and {\em coherence} properties that
describe how the region's data may be used by computations.
\end{itemize}

While data in Legion is organized in a hierarchy of logical region forests, 
computation is organized in a hierarchy of tasks.  Logical regions and tasks interact
through a system of privileges specifying operations that a task
can perform on a logical region argument (read, write, or reduce).  
Tasks can only access logical regions for which they 
possess privileges.  Tasks acquire privileges from their calling task
or by creating new logical regions.  A key restriction is that 
a sub-task can only be passed a subset of the privileges of its parent task.

Through multiple partitions of a logical region, a given datum may
belong to multiple different regions.  Thus, allowing multiple
simultaneous partitions of the same data introduces
aliasing---distinct regions may have overlapping data.\footnote{This
  is why we use the term {\em logical} regions: language-level regions
  name sets of data but do not imply a physical layout or placement in
  the memory hierarchy. A separate system of {\em physical regions} at
  runtime holds copies of data in the language-level logical regions
  \cite{Legion12}.}  We believe permitting very flexible aliasing is
necessary, but it is also presents a major challenge. For example, to
detect that tasks can execute in parallel it is sufficient to prove
that they work on disjoint data.

@@

To efficiently enforce restrictions that guarantee the safety and
correctness of program executions for uses of first-class regions,
their privilege and coherence properties, and arbitrary region
aliasing, the Legion runtime relies on a combination of static and
dynamic checks.  A key theme that runs through the design is that very
hard static analysis problems (e.g., alias analysis) are often both easy and
inexpensive if they can be done dynamically at the granularity of regions instead of individual
heap locations, while checks that must be done on individual heap locations are only
cheap if done statically.  Explicating and evaluating the design of Legion's
static and dynamic semantics and its implications is the main topic of this paper,
which we present as a series of contributions:
\begin{itemize}
\item We present a type system for {\em Core Legion} programs that statically
  verifies the safety of individual pointer references and region
  privileges at task call boundaries (Section~\ref{subsec:coretypes}).
  A subtle issue arises in checking privileges when regions are
  stored in and later retrieved from heap data structures.  The type
  system provides enough static information that the Legion runtime
  can leverage its dynamic information about region aliasing to
  perform dynamic privelege checks guaranteeing safety.

\item We present a parallel operational semantics for Core Legion that
  may be of independent interest.  This semantics is compositional,
  hierarchical, and asynchronous, reflecting the way such programs
  actually execute on the hardware.  However, the semantics also
  directly models conflicting memory operations between tasks in cases of relaxed
  coherence, making explicit the boundary between deterministic and
  possible non-deterministic behavior (Section~\ref{subsec:opsemantics}).

\item We prove the soundness of Legion's privelege system (Section~\ref{sec:soundness}).

\item We use the soundness of privileges to prove the soundness of
  Legion's region coherence modes (Section~\ref{sec:coherence}).

\item We show that if expressions $e_1$ and $e_2$ are {\em non-interfering} (can be executed in parallel), then subexpressions
$e_1'$ of $e_1$ and $e_2'$ of $e_2$ are also non-interfering (Section~\ref{sec:scheduling}).  
This result is the basis
for a hierarchical, distributed scheduler in the Legion implementation which is crucial for high performance
on the target class of machines;  any form of centralized scheduling would be a bottleneck because of the
large communication latencies in the target class of machines.
%; on the target class of machines, any centralized
%scheduler would be a serious bottleneck.

\item We give experimental evidence for the Legion design.  On three
  real-world applications, we show that dynamic region pointer checks
  would be expensive, justifying checking this aspect of the type
  system statically.  We also show that the cost of region aliasing
  checks is low, showing that a much more expressive and dynamic
  language with aliasing is not incompatible with both high
  performance and safety (Section~\ref{sec:evaluation}).

>>>>

Our own experience writing high-performance applications
using the current industry standard mix of MPI, shared-memory
threads, and CUDA, has taught us that a
single, static data partitioning is insufficient.  In practice, the
best way to partition data is often a function of the data
itself---i.e., the partitions must be dynamically computed.
Furthermore, applications often need multiple, simultaneous partitions of
the same data, each providing different views on to that data.

In this work, we present the static and dynamic semantics of Legion \cite{Legion12},
a parallel programming model that efficiently supports multiple, dynamic partitions
of data. Legion's {\em logical regions} are an
abstraction capturing {\em locality} (data that will be used together,
and therefore should be colocated) and {\em independence} (disjoint data
that may be operated on in parallel).  Logical regions have 
properties that allow programmers to communicate information
about the structure of program data to the Legion implementation:
\begin{itemize}
\item  logical regions are first-class values in Legion
and may be dynamically allocated and stored in data structures;

@@

\item {\em privilege} and {\em coherence} properties that
describe how a logical region's data may be used by computations.
\end{itemize}

While data in Legion is organized in a hierarchy of logical regions,
computation is organized in a hierarchy of tasks.  Logical regions and tasks interact
through a system of {\em privileges} specifying operations that a task
can perform on a logical region argument (read, write, or reduce).  
Tasks can only access logical regions for which they 
possess privileges.  
%Tasks acquire privileges from their calling task
%or by creating new logical regions.  A key restriction is that 
%a sub-task can only be passed a subset of the privileges of its parent task.

Through multiple partitions of a logical region, a given datum may
belong to multiple different regions.  Thus, allowing multiple
simultaneous partitions of the same data introduces
{\em aliasing}---distinct regions may have overlapping data.\footnote{Thus
  the term {\em logical} regions: language-level regions
  name sets of data but do not imply a physical layout or placement in
  the memory hierarchy. A separate system of {\em physical regions} at
  runtime holds copies of data in the language-level logical regions
  \cite{Legion12}.}

@@

To efficiently guarantee the safety and
correctness of programs using first-class regions,
privilege and coherence properties, and arbitrary region
aliasing, Legion relies on a combination of static and
dynamic checks.  A key theme that runs through the design is that very
hard static analysis problems (specifically, alias analysis) are often both easy and
inexpensive if they can be done dynamically at the granularity of regions instead of individual
heap locations, while checks that must be done on individual values are only
cheap if done statically.  Explicating and evaluating the design of Legion's
static and dynamic semantics and its implications are the topics of this paper,
which we present as a series of contributions:
\begin{itemize}
\item We present a type system for {\em Core Legion} programs that statically
  verifies the safety of individual pointers and region
  privileges at task call boundaries (Section~\ref{subsec:coretypes}).
%  A subtle issue arises in checking privileges when regions are
%  stored in and later retrieved from heap data structures.  The type
%  system provides enough static information that the Legion runtime
%  can leverage its dynamic information about region aliasing to
%  perform dynamic privelege checks guaranteeing safety.

\item We present a parallel operational semantics for Core Legion that
  may be of independent interest.  This semantics is compositional,
  hierarchical, and asynchronous, reflecting the way such programs
  actually execute on the hardware.  The semantics also
  models conflicting memory operations between tasks in cases of relaxed
  coherence, making explicit the boundary between deterministic and
  possible non-deterministic behavior (Section~\ref{subsec:opsemantics}).

\item We prove the soundness of Legion's static type and privelege system (Section~\ref{sec:soundness}).

\item We use soundness of the type system to prove the soundness of
  Legion's region coherence modes (Section~\ref{sec:coherence}).

\item Again using the sounds of the type system,  we show that if expressions $e_1$ and $e_2$ are {\em non-interfering} (can be executed in parallel), then subexpressions
$e_1'$ of $e_1$ and $e_2'$ of $e_2$ are also non-interfering (Section~\ref{sec:scheduling}).  
This result is the basis
for a hierarchical, distributed scheduler in the Legion implementation which is crucial for high performance
on the target class of machines.
%  any form of centralized scheduling would be a bottleneck because of the
%large communication latencies in the target class of machines.
%; on the target class of machines, any centralized
%scheduler would be a serious bottleneck.

\item We give experimental evidence for the Legion design.  On three
  real-world applications, we show that dynamic region pointer checks
  would be expensive, justifying checking this aspect of the type
  system statically.  We also show that the cost of region aliasing
  checks is low, showing that an expressive and dynamic
  language with aliasing is not incompatible with both high
  performance and safety (Section~\ref{sec:evaluation}).
